---
title: "Hierarchical and Nested Data"
author: "Max Moldovan and Dino Sejdinovic"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical and Nested Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

## Why Hierarchy Matters

Many real-world datasets have nested structure:

- Patients within hospitals
- Students within schools
- Plots within farms
- Repeated measures within subjects

Standard kernel tests assume independent observations. When observations
are clustered, **within-cluster correlation inflates type I error**.
`hierarchical_test()` accounts for this by decomposing the test
statistic and permuting within clusters.

## Example: Agriculture Trial

Imagine a randomised fertiliser trial across 20 farms, each with
30 plots.

```{r agri-example}
library(kernR)
set.seed(42)

n_farms <- 20
n_plots <- 30
n <- n_farms * n_plots
farm_id <- rep(1:n_farms, each = n_plots)

# Farm-level random effects
farm_effect <- rnorm(n_farms, sd = 2)[farm_id]
soil <- matrix(rnorm(n * 2), n, 2)

# Treatment assignment (partially confounded by soil)
treatment <- rbinom(n, 1, plogis(0.3 * soil[, 1]))

# Yield: treatment has a real effect + farm random effect
yield <- 0.8 * treatment + farm_effect + 0.5 * soil[, 1] + rnorm(n)

result <- hierarchical_test(
  y = yield,
  treatment = treatment,
  covariates = soil,
  cluster_id = farm_id,
  method = "dr-date",
  n_permutations = 100,
  weight_method = "icc",
  seed = 1
)
result
```

The test detects the treatment effect while correctly accounting for
the farm-level clustering.

## Decomposition: Within vs Between

The test provides both components:

```{r decomposition}
cat("Within-cluster average statistic:", mean(result$hierarchical$within_stats, na.rm = TRUE), "\n")
cat("Between-cluster statistic:", result$hierarchical$between_stat, "\n")
cat("Combined statistic:", result$statistic, "\n")
cat("Weight method:", result$hierarchical$weight_method, "\n")
```

## Weight Methods

| Method | Behaviour |
|--------|-----------|
| `"equal"` | Equal weight to within and between components |
| `"icc"` | Weight by ICC (more between-weight when clusters differ) |
| `"within_only"` | Ignore between-cluster variation |

## Example: No Treatment Effect

```{r null-hierarchical}
set.seed(42)
yield_null <- farm_effect + 0.5 * soil[, 1] + rnorm(n) # No treatment effect

result_null <- hierarchical_test(
  y = yield_null,
  treatment = treatment,
  covariates = soil,
  cluster_id = farm_id,
  method = "dr-date",
  n_permutations = 100,
  seed = 1
)
cat("P-value under null:", result_null$p_value, "\n")
```

## When to Use Hierarchical Tests

| Scenario | Recommendation |
|----------|---------------|
| Independent observations | Use standard `dr_date_test()` / `bd_hsic_test()` |
| Clustered data (known groups) | Use `hierarchical_test()` with `cluster_id` |
| Few large clusters | `weight_method = "icc"` |
| Many small clusters | `weight_method = "equal"` |
| Unsure about between-cluster effects | `weight_method = "within_only"` (conservative) |

```{r session-info}
sessionInfo()
```
