---
title: "Theory: Kernel Mean Embeddings, Causal Adjustment, and Doubly Robust Tests"
author: "Max Moldovan and Dino Sejdinovic"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory: Kernel Mean Embeddings, Causal Adjustment, and Doubly Robust Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 150
)
```

## 1. Introduction

This vignette is the canonical theory reference for **kernR**. It is a
derivation walk-through, not a presentation of new theory: the goal is
to make the mathematical machinery underlying every test in the
package legible to a reader who already knows some statistics but has
not previously worked with kernel mean embeddings.

We move from the ground up:

1. The reproducing kernel Hilbert space (RKHS) and what it lets us do.
2. The Maximum Mean Discrepancy (MMD), which compares two distributions
   by comparing their embeddings.
3. The Hilbert-Schmidt Independence Criterion (HSIC), which tests
   independence by embedding the joint distribution and the product
   of marginals.
4. The backdoor-adjusted HSIC (bd-HSIC) of Hu, Sejdinovic, and Evans
   (2024), which extends HSIC to causal association under measured
   confounding.
5. Conditional mean embeddings (CMEs) and what they buy us beyond
   estimating means.
6. The doubly robust kernel statistics of Fawkes, Hu, Evans, and
   Sejdinovic (2024) -- DR-DATE and DR-DETT -- for distributional
   treatment effects.
7. kernR's hierarchical extension, which lifts these tests to
   multi-level data with a within/between decomposition and
   cluster-respecting permutation.

Each section pairs the mathematics with one short, runnable kernR
example so the reader can see the formulas land in code. We avoid
computational scalability topics here -- Nystrom approximations and
random Fourier features are addressed in the *performance* vignette,
not this one.

Notation. $\mathcal{X}$, $\mathcal{Y}$, $\mathcal{Z}$ are measurable
spaces. $P, Q$ are probability measures on them. We write
$X \sim P$ and $X' \sim P$ for independent copies. Lower-case
$x_1, \ldots, x_n$ denote a sample; upper-case $X$ denotes the random
variable. $\langle \cdot, \cdot \rangle_{\mathcal{H}}$ is the inner
product in an RKHS $\mathcal{H}$ and $\| \cdot \|_{\mathcal{H}}$ is
its norm.

## 2. Reproducing Kernel Hilbert Spaces

A **kernel** $k : \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ is a
symmetric, positive semi-definite function: for any
$x_1, \ldots, x_n \in \mathcal{X}$ and $c \in \mathbb{R}^n$,
$$
\sum_{i=1}^n \sum_{j=1}^n c_i c_j \, k(x_i, x_j) \ge 0.
$$
By the Moore-Aronszajn theorem, every such $k$ corresponds uniquely
to a Hilbert space of functions $\mathcal{H}_k$ on $\mathcal{X}$,
called its **reproducing kernel Hilbert space**, satisfying:

- $k(\cdot, x) \in \mathcal{H}_k$ for every $x \in \mathcal{X}$;
- the **reproducing property**:
  $f(x) = \langle f, k(\cdot, x) \rangle_{\mathcal{H}_k}$ for all
  $f \in \mathcal{H}_k$, $x \in \mathcal{X}$.

Define the **canonical feature map** $\phi(x) := k(\cdot, x)$. Then
$$
k(x, x') = \langle \phi(x), \phi(x') \rangle_{\mathcal{H}_k},
$$
which is the *kernel trick*: an inner product in a potentially
infinite-dimensional feature space is computed by evaluating the
kernel.

For most of the package we use the **radial basis function (RBF)**
kernel,
$$
k_\sigma(x, x') = \exp\!\left(-\frac{\|x - x'\|^2}{2 \sigma^2}\right),
$$
with bandwidth $\sigma > 0$. The RBF kernel is **characteristic** on
$\mathbb{R}^d$: the map $P \mapsto \mathbb{E}_{X \sim P} \phi(X)$ from
probability measures to $\mathcal{H}_k$ is injective. Two distributions
have the same mean embedding if and only if they are equal. Without
this property, the tests we develop below would be unable to distinguish
some pairs of distributions and would silently lose power.

kernR exposes the kernel as a first-class object via `kernel_spec()`.
Bandwidth selection defaults to the **median heuristic**, $\sigma =
\mathrm{median}(\|x_i - x_j\|)$, which is robust and rarely wrong by
much. Other supported families are Matern (smoothness controlled by
$\nu$), linear, and polynomial.

```{r rkhs-demo}
library(kernR)

set.seed(2026)
x <- matrix(rnorm(20), 10, 2)

# Default RBF with median heuristic
k <- kernel_spec()
K <- kernel_matrix(x, kernel = k)

# Symmetry and PSD
all.equal(K, t(K))
ev <- eigen(K, symmetric = TRUE, only.values = TRUE)$values
all(ev > -1e-8)
round(range(ev), 4)
```

The Gram matrix is symmetric and (numerically) PSD, as it must be.

## 3. Maximum Mean Discrepancy (MMD)

Suppose we have samples from two distributions $P$ on $\mathcal{X}$
and $Q$ on $\mathcal{X}$ and want to test $H_0 : P = Q$. The
**kernel mean embedding** of $P$ in $\mathcal{H}_k$ is
$$
\mu_P := \mathbb{E}_{X \sim P}[\phi(X)] = \mathbb{E}_{X \sim P}[k(\cdot, X)].
$$
The Maximum Mean Discrepancy is the distance between embeddings,
$$
\mathrm{MMD}(P, Q; \mathcal{H}_k) = \| \mu_P - \mu_Q \|_{\mathcal{H}_k}.
$$
Squaring and using bilinearity of the inner product gives the form
that drives all sample estimators:
$$
\mathrm{MMD}^2(P, Q) = \mathbb{E}[k(X, X')] + \mathbb{E}[k(Y, Y')]
  - 2\, \mathbb{E}[k(X, Y)],
$$
with $X, X' \overset{\text{iid}}{\sim} P$,
$Y, Y' \overset{\text{iid}}{\sim} Q$, all four independent.

When $k$ is characteristic, $\mathrm{MMD}^2(P, Q) = 0 \iff P = Q$
(Gretton et al., 2012). The squared MMD therefore behaves as a proper
divergence between distributions on $\mathcal{H}_k$.

### 3.1 Sample estimators

Given $x_1, \ldots, x_m \sim P$ and $y_1, \ldots, y_n \sim Q$, write
$K^{xx}_{ij} = k(x_i, x_j)$, $K^{yy}_{ij} = k(y_i, y_j)$,
$K^{xy}_{ij} = k(x_i, y_j)$. The **biased V-statistic** is
$$
\widehat{\mathrm{MMD}}^2_b
  = \frac{1}{m^2}\sum_{i,j} K^{xx}_{ij}
  + \frac{1}{n^2}\sum_{i,j} K^{yy}_{ij}
  - \frac{2}{mn}\sum_{i,j} K^{xy}_{ij},
$$
which includes the diagonal terms ($i = j$). The **unbiased
U-statistic** removes them:
$$
\widehat{\mathrm{MMD}}^2_u
  = \frac{1}{m(m-1)}\sum_{i \ne j} K^{xx}_{ij}
  + \frac{1}{n(n-1)}\sum_{i \ne j} K^{yy}_{ij}
  - \frac{2}{mn}\sum_{i,j} K^{xy}_{ij}.
$$
The unbiased form is preferred for inference because its expectation
under $H_0$ is exactly zero. Under $H_0$, both estimators are
degenerate U/V-statistics whose limiting distributions are mixtures
of $\chi^2_1$ variables; closed-form thresholds exist but are awkward
in practice. kernR uses **permutation inference** instead -- shuffle
the pooled labels, recompute the statistic, and compare to the
observed value.

```{r mmd-demo}
set.seed(7)
m <- n <- 100

# Same distribution: MMD should be small
x <- matrix(rnorm(m * 2), m, 2)
y <- matrix(rnorm(n * 2), n, 2)
res_null <- mmd_test(x, y, n_permutations = 200, seed = 1)
res_null

# Mean shift: MMD should be large
y_shift <- matrix(rnorm(n * 2, mean = 0.5), n, 2)
res_alt <- mmd_test(x, y_shift, n_permutations = 200, seed = 1)
res_alt
```

The first p-value is large; the second is small. This is the basic
two-sample test. Everything that follows builds on it.

## 4. Hilbert-Schmidt Independence Criterion (HSIC)

To test independence of $X$ and $Y$ -- $H_0 : P_{XY} = P_X \otimes P_Y$
-- we apply the same idea to the joint and product distributions on
$\mathcal{X} \times \mathcal{Y}$. With kernels $k$ on $\mathcal{X}$
and $\ell$ on $\mathcal{Y}$, the product kernel is
$k((x_1, y_1), (x_2, y_2)) = k(x_1, x_2) \ell(y_1, y_2)$, and
$$
\mathrm{HSIC}(X, Y; k, \ell)
  = \big\| \mu_{P_{XY}} - \mu_{P_X \otimes P_Y} \big\|^2_{\mathcal{H}_k \otimes \mathcal{H}_\ell}.
$$
Equivalently, HSIC is the squared Hilbert-Schmidt norm of the
cross-covariance operator $C_{XY} : \mathcal{H}_\ell \to \mathcal{H}_k$
defined by
$$
\langle f, C_{XY} g \rangle_{\mathcal{H}_k}
  = \mathrm{Cov}\!\big(f(X), g(Y)\big),
\qquad f \in \mathcal{H}_k, \; g \in \mathcal{H}_\ell,
$$
giving the population identity
$$
\mathrm{HSIC}(X, Y) = \| C_{XY} \|_{\mathrm{HS}}^2.
$$
When $k$ and $\ell$ are characteristic (equivalently, when the product
kernel is characteristic on $\mathcal{X} \times \mathcal{Y}$),
$\mathrm{HSIC}(X, Y) = 0 \iff X \perp Y$ (Gretton et al., 2005). This
is what makes HSIC a *non-linear* independence test: it sees any
dependence detectable through the chosen RKHS, not just linear
correlation.

### 4.1 Sample formula via centring

Given $\{(x_i, y_i)\}_{i=1}^n$, form Gram matrices $K, L \in
\mathbb{R}^{n \times n}$ with $K_{ij} = k(x_i, x_j)$, $L_{ij} = \ell(y_i, y_j)$,
and the centring matrix
$$
H = I_n - \tfrac{1}{n} \mathbf{1}_n \mathbf{1}_n^\top,
$$
which projects onto the orthogonal complement of the constant
direction. The biased V-statistic estimator is
$$
\widehat{\mathrm{HSIC}}(X, Y) = \frac{1}{n^2}\,\mathrm{tr}(K H L H).
$$
The trick is that $K H L H$ pre- and post-multiplies by the
centring matrix, which is exactly the operation that subtracts row
and column means from $K$ and $L$. Because trace is invariant under
cyclic permutation, an equivalent and slightly cheaper form is
$\mathrm{tr}((HKH)(HLH)) / n^2$. An unbiased U-statistic version
exists (Song et al., 2012) and is also implemented; the choice
rarely matters under permutation inference because the null
distribution is generated by relabelling, not by appeal to a
limiting distribution.

```{r hsic-derivation}
set.seed(11)
n <- 200
xv <- rnorm(n)
yv <- xv^2 + rnorm(n, sd = 0.3)   # nonlinear, zero linear correlation

# Manual HSIC from first principles
K <- kernel_matrix(xv)
L <- kernel_matrix(yv)
H <- diag(n) - matrix(1 / n, n, n)
hsic_manual <- sum(diag(K %*% H %*% L %*% H)) / n^2
hsic_manual

# kernR's test recovers the same statistic and adds a permutation p-value
res_hsic <- hsic_test(xv, yv, n_permutations = 200, seed = 1)
res_hsic
round(c(manual = hsic_manual, package = res_hsic$statistic), 6)
cat("Pearson correlation: ", round(cor(xv, yv), 3), "\n")
```

The manual calculation matches `hsic_test()` to numerical precision,
even though the Pearson correlation is essentially zero. That is
exactly the case HSIC is designed for.

## 5. Backdoor-Adjusted HSIC (bd-HSIC)

HSIC tests *associational* independence. Suppose instead we want a
**causal** statement -- "does intervening on $X$ change the
distribution of $Y$?" -- and we have measured confounders
$Z$ that satisfy Pearl's backdoor criterion (Pearl, 2009): every
back-door path from $X$ to $Y$ is blocked by $Z$, and $Z$ contains
no descendants of $X$.

Under that DAG assumption, the interventional distribution factorises
as
$$
p\!\big(y \mid \mathrm{do}(x)\big)
  = \int p(y \mid x, z)\, p(z)\, dz,
$$
which is **not** the conditional $p(y \mid x)$ used by naive HSIC.
The two coincide only when $X \perp Z$, i.e., when there is no
confounding to begin with.

Hu, Sejdinovic, and Evans (2024) note that the do-null hypothesis
$$
H_0 :\; p\!\big(y \mid \mathrm{do}(x)\big) = p^*(y) \quad \text{for all } x,
$$
where $p^*(y) := \int p(y \mid x, z)\, p(z)\, p(x)\, dz\, dx$ is
the marginal interventional outcome distribution, is equivalent to a
weighted independence statement on the observational distribution:
$$
\mathrm{HSIC}_w(X, Y) := \big\| C_{XY}^w \big\|_{\mathrm{HS}}^2 = 0,
$$
with weights
$$
w(x, z) = \frac{p(x)}{p(x \mid z)},
$$
the inverse of the conditional density of $X$ given $Z$ relative to
the marginal. Intuitively: reweighting observational $(X, Y, Z)$
samples by $w(x, z)$ breaks the dependence between $X$ and $Z$,
which is precisely what intervention does.

### 5.1 Estimating the density ratio by NCE

The weight $w$ is a ratio of densities and is hard to estimate by
plug-in. Hu et al. recast it as a binary classification problem
known as **noise contrastive estimation** (NCE) or *density ratio
by classification*. Form a labelled dataset:

- positives ($u = 1$): observed $(x_i, z_i)$ pairs from $p(x, z)$;
- negatives ($u = 0$): synthetic pairs $(x'_j, z_j)$ where $x'_j$ is
  drawn from the marginal $p(x)$ independently of $z_j$, so
  $(x'_j, z_j) \sim p(x) p(z)$.

A classifier $\widehat{r}(x, z) = \Pr(u = 1 \mid x, z)$ trained on this
mixture yields a density ratio via
$$
\frac{p(x, z)}{p(x) p(z)}
  = \frac{p(x \mid z)}{p(x)}
  = \frac{\widehat{r}(x, z)}{1 - \widehat{r}(x, z)},
$$
so that the bd-HSIC weight is its reciprocal,
$$
\widehat{w}(x, z) = \frac{1 - \widehat{r}(x, z)}{\widehat{r}(x, z)}.
$$
Logistic regression (`density_ratio = "logistic"`), random forests
(`"ranger"`), or gradient boosting (`"xgboost"`) are all valid
classifiers; choice depends on the smoothness, dimension, and
sample size. `"rulsif"` (Yamada et al., 2013) provides an
alternative direct-fitting estimator that is more robust under
limited overlap.

### 5.2 Weighted HSIC and cluster permutation

The sample bd-HSIC plugs the weights into a centred, weighted
trace:
$$
\widehat{\mathrm{HSIC}}_w(X, Y)
  = \frac{1}{n^2} \mathrm{tr}(W K H L H),
$$
where $W = \mathrm{diag}(w_1, \ldots, w_n)$ with $w_i =
\widehat{w}(x_i, z_i)$. (Several equivalent symmetric forms exist;
the package uses one that preserves PSD-ness for the permutation
mechanism.)

Inference is by **cluster-based permutation**. Naively shuffling $Y$
breaks not only the $X$-$Y$ link but also the $Z$-$Y$ link, which
inflates Type I error under confounding. Instead, kernR clusters
observations into groups with similar conditional densities
$p(x \mid z)$ -- typically by binning $\widehat{r}(x, z)$ -- and
permutes $Y$ only *within* clusters. This preserves the confounding
structure while breaking causal association, yielding a calibrated
null.

```{r bdhsic-demo}
set.seed(13)
n <- 250
zv <- matrix(rnorm(n * 2), n, 2)
xv <- 0.5 * zv[, 1] + rnorm(n)            # X depends on Z
yv <- 0.5 * zv[, 1] + zv[, 2] + rnorm(n)  # Y depends on Z, NOT on X

# Naive HSIC sees confounded association
res_naive <- hsic_test(xv, yv, n_permutations = 200, seed = 1)
cat("Naive HSIC p-value (confounding only):", round(res_naive$p_value, 3), "\n")

# bd-HSIC correctly reports no causal effect
res_bd <- bd_hsic_test(xv, yv, zv, n_permutations = 200, seed = 1)
res_bd
```

The naive HSIC mislabels confounded association as causal; bd-HSIC
correctly fails to reject. This is the central benefit of the
backdoor adjustment.

## 6. Conditional Mean Embeddings and Distributional Effects

Mean embeddings $\mu_P$ are unconditional. To talk about treatment
effects on the *distribution* of $Y$ given covariates $X$, we need
the **conditional mean embedding** (Song, Huang, Smola, & Fukumizu,
2009):
$$
\mu_{Y \mid X = x} := \mathbb{E}\!\big[\phi(Y) \mid X = x\big].
$$
The CME is a function $x \mapsto \mu_{Y \mid X = x}$ valued in
$\mathcal{H}_\ell$. Under standard regularity, it is the solution
to a kernel ridge regression of $\phi(Y)$ on $\phi(X)$:
$$
\widehat{\mu}_{Y \mid X = x}
  = \mathbf{l}(\cdot)^\top (K_X + n \lambda I)^{-1} \mathbf{k}_X(x),
$$
where $\mathbf{k}_X(x)$ is the $n$-vector of $k(x_i, x)$, $K_X$ is
the Gram matrix on $\{x_i\}$, $\mathbf{l}(\cdot) = (\ell(\cdot, y_1),
\ldots, \ell(\cdot, y_n))^\top$, and $\lambda > 0$ is a ridge
parameter.

CMEs let us encode the *distribution* of potential outcomes
$Y(t) \mid X$ as a function of $X$, then test or contrast those
distributions in a Hilbert space. Crucially, distributional effects
include changes in variance, skewness, modality, or tail
behaviour -- not just the mean. Standard causal estimators (Double
ML, TMLE) target the mean and have no power against pure variance
or shape effects.

```{r cme-intuition}
set.seed(17)
n <- 400
tt <- rbinom(n, 1, 0.5)
# Identical mean, doubled variance under treatment
yv <- ifelse(tt == 1, rnorm(n, 0, 2), rnorm(n, 0, 1))
c(mean_diff = mean(yv[tt == 1]) - mean(yv[tt == 0]),
  sd_treated = sd(yv[tt == 1]),
  sd_control = sd(yv[tt == 0]))

# MMD picks up the distributional difference even though means coincide
res <- mmd_test(yv[tt == 1], yv[tt == 0], n_permutations = 200, seed = 1)
res$p_value
```

Means agree; the distributions do not; MMD detects the difference.
This is the regime DR-DATE and DR-DETT formalise causally.

## 7. Doubly Robust Distributional Tests

Fawkes, Hu, Evans, and Sejdinovic (2024) construct distributional
analogues of doubly robust ATE estimators. The setting is the
standard binary-treatment causal model: treatment $T \in \{0, 1\}$,
covariates $X$, outcome $Y$, with potential outcomes $Y(0)$ and
$Y(1)$, observed outcome $Y = T \cdot Y(1) + (1-T) \cdot Y(0)$, and
the usual identification assumptions:

- **unconfoundedness**: $\{Y(0), Y(1)\} \perp T \mid X$;
- **overlap**: $0 < e(x) := \Pr(T = 1 \mid X = x) < 1$;
- **SUTVA**.

We want to test $H_0 : P_{Y(1)} = P_{Y(0)}$ (DR-DATE) or, more
modestly, $H_0 : P_{Y(1) \mid T = 1} = P_{Y(0) \mid T = 1}$ (DR-DETT).

### 7.1 Building blocks

Two nuisance models are required:

- a **propensity** model $\widehat{e}(x)$ for $\Pr(T = 1 \mid X = x)$
  (logistic, ranger, or xgboost, with **cross-fitting** to avoid
  bias from in-sample reuse);
- an **outcome embedding** model
  $\widehat{m}_t(x) = \widehat{\mu}_{Y \mid X = x, T = t}$, the
  CME of $Y$ given $X$ within each treatment arm.

### 7.2 The DR-DATE statistic

Let $\widehat{m}_1$ and $\widehat{m}_0$ be the fitted CMEs under
treatment and control. The doubly robust estimator of $\mu_{Y(t)}$ is
$$
\widehat{\mu}_{Y(1)}
  = \frac{1}{n} \sum_{i=1}^n \left[
    \frac{T_i}{\widehat{e}(X_i)}\big(\phi(Y_i) - \widehat{m}_1(X_i)\big)
    + \widehat{m}_1(X_i)
  \right],
$$
and symmetrically for $\widehat{\mu}_{Y(0)}$ with $T \to 1 - T$ and
$\widehat{e} \to 1 - \widehat{e}$. The DR-DATE statistic is
$$
\widehat{T}_{\text{DR-DATE}}
  = \big\| \widehat{\mu}_{Y(1)} - \widehat{\mu}_{Y(0)} \big\|_{\mathcal{H}_\ell}^2,
$$
i.e., a squared MMD between two doubly robust embedding estimators.

### 7.3 Why double robustness

Each summand has the **augmented inverse-probability-weighted (AIPW)**
form. Standard AIPW algebra extends to embeddings: the estimator is
consistent for the true $\mu_{Y(t)}$ if either

- $\widehat{e}$ is consistent for the propensity, **or**
- $\widehat{m}_t$ is consistent for the conditional mean embedding,

without requiring both. A misspecification in one nuisance is absorbed
provided the other is correct. Asymptotic normality (in $\mathcal{H}_\ell$)
holds under cross-fitting at typical sub-parametric rates, with the
familiar product-rate condition: the rate of $\|\widehat{\mu}_{Y(t)} -
\mu_{Y(t)}\|$ is governed by the product of nuisance estimation
errors. Permutation tests under a calibrated **bin-based** permutation
(stratifying by propensity quantiles, mirroring the cluster
permutation of bd-HSIC) recover correct Type I error in finite
samples.

### 7.4 DR-DETT: effect on the treated

DR-DETT targets the treated subpopulation, requiring only one-sided
overlap ($e(x) < 1$ on the support of treated units). The construction
uses
$$
\widehat{\mu}_{Y(0) \mid T = 1}
  = \frac{1}{n_1} \sum_{i : T_i = 1}
    \widehat{m}_0(X_i)
    + \frac{1}{n_1} \sum_{i : T_i = 0}
    \frac{\widehat{e}(X_i)}{1 - \widehat{e}(X_i)}
    \big(\phi(Y_i) - \widehat{m}_0(X_i)\big),
$$
which is the embedding analogue of the standard DETT/ATT influence
function. The treated arm contributes $\phi(Y_i)$ directly; control
units are upweighted by the odds ratio $e/(1-e)$ to estimate the
counterfactual outcome embedding under no treatment.

```{r drdate-demo}
set.seed(19)
n <- 300
xv <- matrix(rnorm(n * 2), n, 2)
tv <- rbinom(n, 1, plogis(0.3 * xv[, 1]))
# Variance effect only -- mean shift is essentially zero
yv <- (1 - tv) * rnorm(n, 0, 1) + tv * rnorm(n, 0, 2.5) + 0.4 * xv[, 1]

res_date <- dr_date_test(yv, tv, xv,
  propensity_model = "logistic",
  outcome_model = "krr",
  n_permutations = 150,
  seed = 1
)
res_date
```

DR-DATE detects a distributional difference even when the mean is
unchanged -- a test that mean-targeted estimators cannot construct
without ad-hoc transformations.

## 8. Hierarchical Extension

Real applications -- patients within hospitals, plots within fields,
students within schools -- violate the i.i.d. assumption underlying
all the tests above. Two failure modes appear:

1. **Inflated Type I error.** Permuting individuals across clusters
   ignores within-cluster dependence; the permutation null
   distribution becomes too narrow.
2. **Confounded between-cluster comparisons.** Differences between
   clusters can drive a "treatment effect" that, conditional on the
   cluster, is absent.

kernR's `hierarchical_test()` decomposes the kernel statistic into
a within-cluster and a between-cluster contribution and tests
each in isolation, then combines them with a configurable
weighting. Concretely, for a kernel statistic $S$ on data
$\{(x_{cj}, y_{cj}) : c = 1, \ldots, C, j = 1, \ldots, n_c\}$ with
cluster id $c$, write
$$
S_{\text{total}}
  = S_{\text{within}} + S_{\text{between}},
$$
mirroring the variance decomposition implicit in random-effects
models. The intra-class correlation
$$
\rho = \frac{\sigma^2_{\text{between}}}{\sigma^2_{\text{between}} + \sigma^2_{\text{within}}}
$$
is estimated and used to set weights when `weight_method = "icc"`,
following the principle that high-$\rho$ data should down-weight
within-cluster signal (which is dominated by cluster-mean variation).
`weight_method = "equal"` and `weight_method = "within_only"` provide
explicit alternatives -- the latter is appropriate when betweens
are known to be confounded.

The permutation engine is **cluster-respecting**: under the
null, we permute *cluster labels* (between-cluster permutation) or
*outcomes within clusters* (within-cluster permutation), never
across clusters. This preserves the dependence structure that
generated the data.

```{r hierarchical-demo, eval = FALSE}
# Marked eval = FALSE only because hierarchical permutation is the
# heaviest path; the call shape is correct and stable.
set.seed(23)
C <- 20            # clusters
nc <- 15           # individuals per cluster
n <- C * nc
cluster_id <- rep(seq_len(C), each = nc)
u <- rnorm(C, sd = 0.7)[cluster_id]   # cluster random effect
xv <- matrix(rnorm(n * 2), n, 2)
tv <- rbinom(n, 1, plogis(0.3 * xv[, 1] + u))
yv <- 0.6 * tv + 0.4 * xv[, 1] + u + rnorm(n)

res_h <- hierarchical_test(
  y = yv, treatment = tv, covariates = xv,
  cluster_id = cluster_id,
  method = "dr-date",
  weight_method = "icc",
  n_permutations = 200, seed = 1
)
res_h
```

This recovers the validity of distributional tests under the
multi-level structure that pervades health, agriculture, and
educational data.

## 9. References

Fawkes, J., Hu, R., Evans, R. J., & Sejdinovic, D. (2024). Doubly
robust kernel statistics for testing distributional treatment
effects. *Transactions on Machine Learning Research*.

Gretton, A., Borgwardt, K. M., Rasch, M. J., Schölkopf, B., & Smola,
A. (2012). A kernel two-sample test. *Journal of Machine Learning
Research*, *13*, 723-773.

Gretton, A., Bousquet, O., Smola, A., & Schölkopf, B. (2005).
Measuring statistical dependence with Hilbert-Schmidt norms. In
*Algorithmic Learning Theory* (pp. 63-77). Springer.

Hu, R., Sejdinovic, D., & Evans, R. J. (2024). A kernel test for
causal association via noise contrastive backdoor adjustment.
*Journal of Machine Learning Research*, *25*(160), 1-56.

Pearl, J. (2009). *Causality: Models, reasoning, and inference*
(2nd ed.). Cambridge University Press.

Song, L., Huang, J., Smola, A., & Fukumizu, K. (2009). Hilbert
space embeddings of conditional distributions with applications to
dynamical systems. In *Proceedings of the 26th International
Conference on Machine Learning* (pp. 961-968).

Song, L., Smola, A., Gretton, A., Bedo, J., & Borgwardt, K. (2012).
Feature selection via dependence maximization. *Journal of Machine
Learning Research*, *13*, 1393-1434.

Yamada, M., Suzuki, T., Kanamori, T., Hachiya, H., & Sugiyama, M.
(2013). Relative density-ratio estimation for robust distribution
comparison. *Neural Computation*, *25*(5), 1324-1370.

```{r session-info}
sessionInfo()
```
