---
title: "Performance: Approximations, Benchmarks, and Tuning Guidance"
author: "Max Moldovan and Dino Sejdinovic"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance: Approximations, Benchmarks, and Tuning Guidance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 150
)
```

```{r libs}
library(kernR)
```

## 1. Why approximations

The exact kernel matrix is $n \times n$. Forming it and the operations
that follow scale as $\mathcal{O}(n^2)$ in memory and $\mathcal{O}(n^2 d)$
in compute (for $d$-dimensional inputs). Past a few thousand
observations this becomes the binding constraint, particularly when
permutation inference is layered on top.

kernR exposes two opt-in approximations that target this bottleneck:

- **Nystrom** -- pick $m \ll n$ landmark points, compute the cross block
  $C \in \mathbb{R}^{n \times m}$ and the inner block
  $W \in \mathbb{R}^{m \times m}$, and reconstruct
  $\tilde K = C W^{+} C^{\top}$ via a thresholded eigendecomposition.
  Returns a low-rank factor $U \in \mathbb{R}^{n \times r}$ ($r \le m$)
  with $\tilde K = U U^{\top}$.

- **Random Fourier Features (RFF)** -- for translation-invariant
  kernels (RBF, Matern), Bochner's theorem gives a non-negative
  spectral measure. Sampling $D$ frequencies from it and forming
  $(\cos, \sin)$ features yields a $2D$-dimensional explicit feature
  map $z(x)$ such that $\tilde K(x, x') = z(x)^{\top} z(x')$.

Both surface through a single argument:

```{r api-tour}
kernel_spec("rbf", approx = "nystrom", approx_rank = 100, approx_seed = 1)
kernel_spec("rbf", approx = "rff",     approx_rank = 200, approx_seed = 1)
```

Set `approx_rank = NULL` (the default) and kernR picks
$\max(50, \lceil 0.1 n \rceil)$, capped at $n$.

## 2. Approximation accuracy versus rank

Holding $n$ fixed, both approximations converge to the exact kernel as
their rank parameter grows. The error decay rates differ:

```{r accuracy-curve, fig.cap = "Relative Frobenius error of the approximated kernel matrix versus rank, for an RBF kernel with bandwidth = 1 on a 200-row Gaussian sample (d = 4).", fig.alt = "Log-log plot of relative Frobenius error against rank for Nystrom and RFF approximations of an RBF kernel matrix; both curves decrease as rank grows, with Nystrom (filled circles, solid line) decreasing faster than RFF (filled triangles, dashed line)."}
set.seed(2026)
n <- 200
x <- matrix(rnorm(n * 4), n, 4)
k_exact <- kernel_spec("rbf", bandwidth = 1)
K <- kernel_matrix(x, kernel = k_exact)

ranks <- c(20, 50, 100, 150, 200, 400, 800)
relfro <- function(A, B) sqrt(sum((A - B)^2) / sum(A^2))

results <- data.frame(
  rank = integer(0),
  method = character(0),
  rel_error = numeric(0)
)
for (r in ranks[ranks <= n]) {
  k_nys <- kernel_spec("rbf", bandwidth = 1, approx = "nystrom",
                       approx_rank = r, approx_seed = 1)
  K_nys <- kernel_matrix(x, kernel = k_nys)
  results <- rbind(results, data.frame(
    rank = r, method = "nystrom", rel_error = relfro(K, K_nys)
  ))
}
for (D in ranks) {
  k_rff <- kernel_spec("rbf", bandwidth = 1, approx = "rff",
                       approx_rank = D, approx_seed = 1)
  K_rff <- kernel_matrix(x, kernel = k_rff)
  results <- rbind(results, data.frame(
    rank = D, method = "rff", rel_error = relfro(K, K_rff)
  ))
}

with(results, plot(
  rank, rel_error, type = "n",
  log = "xy", xlab = "rank (m for Nystrom, D for RFF)",
  ylab = "relative Frobenius error",
  main = "Approximation accuracy vs rank"
))
for (m in unique(results$method)) {
  sub <- results[results$method == m, ]
  lines(sub$rank, sub$rel_error, type = "b",
        pch = if (m == "nystrom") 16 else 17,
        lty = if (m == "nystrom") 1 else 2,
        col = if (m == "nystrom") "#1F3A5F" else "#B8420A")
}
legend("topright",
       legend = c("Nystrom", "RFF"),
       col = c("#1F3A5F", "#B8420A"),
       pch = c(16, 17), lty = c(1, 2), bty = "n")
```

Two empirical patterns are visible:

1. **Nystrom converges faster per unit of rank.** Once $m \approx n$
   the reconstruction is exact up to numerical noise from the
   thresholded eigendecomposition.
2. **RFF needs more frequencies.** $D = 200$ is comparable to Nystrom
   $m \approx 30$ in error terms here. The reason is that RFF's basis
   is data-agnostic (random frequencies), whereas Nystrom's columns
   land on actual data points -- so Nystrom captures "where the data
   is" almost for free.

For practical rules of thumb:

- Pick **Nystrom** when memory permits storing $m$ data points twice
  and you can spare the $\mathcal{O}(m^3)$ eigendecomposition. Default
  $m \ge 0.1 n$ is a reasonable starting point.
- Pick **RFF** when you want a *fixed-dimensional* feature map that is
  independent of $n$ (e.g. for streaming or out-of-sample evaluation),
  or when $n$ is so large that storing $n \times m$ is itself a
  problem.

## 3. Statistic-level parity

Kernel approximations only matter for users to the extent that they
preserve the *test statistic*. The next chunk shows that for a moderate
rank both Nystrom and RFF land on the same MMD statistic as the exact
kernel.

```{r stat-parity}
set.seed(42)
n <- 200
x <- matrix(rnorm(n * 2), n, 2)
y <- matrix(rnorm(n * 2, mean = 0.4), n, 2)

k_exact <- kernel_spec("rbf")
k_nys   <- kernel_spec("rbf", approx = "nystrom",
                       approx_rank = 100, approx_seed = 1)
k_rff   <- kernel_spec("rbf", approx = "rff",
                       approx_rank = 1500, approx_seed = 1)

stat_exact <- mmd_test(x, y, kernel = k_exact,
                       n_permutations = 100L, seed = 1)$statistic
stat_nys   <- mmd_test(x, y, kernel = k_nys,
                       n_permutations = 100L, seed = 1)$statistic
stat_rff   <- mmd_test(x, y, kernel = k_rff,
                       n_permutations = 100L, seed = 1)$statistic

c(exact = stat_exact, nystrom = stat_nys, rff = stat_rff)
```

Both approximations sit within a few percent of the exact statistic --
inside the Monte Carlo noise of the permutation null distribution.

## 4. Wall-clock benchmarks

The point of an approximation is speed. The next benchmark exercises
the kernel-matrix-formation step at increasing $n$. We hold the
approximation rank at $0.1 n$ (the default heuristic).

```{r bench-setup}
maybe_bench <- requireNamespace("bench", quietly = TRUE)
```

```{r bench-kernel, eval = maybe_bench}
ns <- c(200, 500, 1000)
out <- lapply(ns, function(n) {
  x <- matrix(rnorm(n * 2), n, 2)
  k_exact <- kernel_spec("rbf")
  k_nys   <- kernel_spec("rbf", approx = "nystrom",
                         approx_rank = max(50L, as.integer(0.1 * n)),
                         approx_seed = 1)
  k_rff   <- kernel_spec("rbf", approx = "rff",
                         approx_rank = max(50L, as.integer(0.1 * n)),
                         approx_seed = 1)
  bench::mark(
    exact   = kernel_matrix(x, kernel = k_exact),
    nystrom = kernel_matrix(x, kernel = k_nys),
    rff     = kernel_matrix(x, kernel = k_rff),
    iterations = 5L,
    check = FALSE,
    relative = FALSE
  )[, c("expression", "median", "mem_alloc")]
})
names(out) <- paste0("n=", ns)
out
```

A full benchmark up to $n = 5000$ is provided as a non-evaluated
chunk -- run it on your own hardware to gauge the trade-off:

```{r bench-large, eval = FALSE}
ns_large <- c(2000, 5000)
for (n in ns_large) {
  x <- matrix(rnorm(n * 2), n, 2)
  k_exact <- kernel_spec("rbf")
  k_nys   <- kernel_spec("rbf", approx = "nystrom",
                         approx_rank = as.integer(0.1 * n), approx_seed = 1)
  k_rff   <- kernel_spec("rbf", approx = "rff",
                         approx_rank = as.integer(0.1 * n), approx_seed = 1)
  print(n)
  print(bench::mark(
    exact   = kernel_matrix(x, kernel = k_exact),
    nystrom = kernel_matrix(x, kernel = k_nys),
    rff     = kernel_matrix(x, kernel = k_rff),
    iterations = 3L, check = FALSE, relative = TRUE
  ))
}
```

## 5. When the approximation does **not** help

Two regimes where the current implementation buys you very little:

1. **Reconstructing the full $n \times n$ matrix.** kernR's
   `kernel_matrix(approx = ...)` materialises the full Gram matrix to
   keep the API uniform. The asymptotic speedup of low-rank
   approximations only realises when downstream code consumes the
   factor directly. That code path is exposed by [`nystrom_factor()`]
   and [`rff_features()`]; HSIC / MMD / DR-tests do not yet exploit it.
2. **Asymmetric calls.** `kernel_matrix(x, y, ...)` with $y \ne x$
   always falls back to exact computation. Cross-blocks are not
   well-posed under either Nystrom (which assumes a square Gram block)
   or RFF without re-aligning random features.

These are limitations of the present cut, not of the underlying
methods. Both will be revisited in a later release.

## 6. Tuning checklist

When you suspect the kernel matrix is your bottleneck:

1. Time the exact path: `system.time(kernel_matrix(x))`.
2. Try `approx = "nystrom"` with the default rank. Compare statistics
   (not p-values) against exact on a small subsample.
3. If the relative error in the statistic is larger than you can
   tolerate, double the rank. Stop when the marginal accuracy gain is
   small.
4. If memory is binding before compute, switch to `approx = "rff"`:
   the feature representation is $n \times 2D$ regardless of $n$.
5. For Matern kernels with small $\nu$ (rough sample paths), prefer
   Nystrom -- the Student-t spectral measure makes RFF slow to
   converge for $\nu \le 1$.

## 7. References

Drineas, P., & Mahoney, M. W. (2005). On the Nystrom method for
approximating a Gram matrix for improved kernel-based learning.
*Journal of Machine Learning Research*, 6, 2153--2175.

Rahimi, A., & Recht, B. (2007). Random features for large-scale kernel
machines. *Advances in Neural Information Processing Systems*, 20.

Sutherland, D. J., & Schneider, J. (2015). On the error of random
Fourier features. *UAI*.
